Live Functional Programming with Typed Holes

### 2018-04-15 calculus.tex
nit

As with the type system for the external language (above), 
we specify expansion bidirectionally \cite{DBLP:conf/ppdp/FerreiraP14}.
%
The synthetic expansion judgement~$\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$ synthesizes a type~$\htau$ from~$\hexp$.
%
Synthetic expansion produces an internal expression~$d$ and a hole context~$\hDelta$.
%
%We say more about hole contexts, which are used in the type assignment judgement, $\hasType{\Delta}{\hGamma}{d}{\htau}$, below.
We describe hole contexts, which serve as ``inputs'' to the type assignment judgement~$\hasType{\Delta}{\hGamma}{d}{\htau}$, further below. 
%
The analytic expansion judgement~$\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$, checks~$\hexp$ against~$\htau$ and produces an expansion~$d$ of type~$\htau'$, and a hole context~$\hDelta$.
%
The governing theorem below establishes that expansions are well-typed and in the analytic case that the type $\htau'$ is consistent with $\htau$.
%
\begin{thm}[Typed Expansion]\label{thm:typed-expansion} ~
  \begin{enumerate}[nolistsep]
    \item
      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
    \item
      If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      then $\tconsistent{\htau}{\htau'}$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$.
  \end{enumerate}
\end{thm}
\noindent

%The reason analytic expansion produces an expansion of consistent
%type is because the subsumption rule, as previously discussed, allows
%us to check an external expression against any type consistent with
%the type the expression actually synthesizes, whereas every internal
%expression can be assigned at most one type, i.e. the following
%standard unicity property holds of the type assignment system.
%
Analytic expansion produces an expansion of consistent type:
%
Intuitively, the subsumption rule, as previously discussed, allows us
to check an external expression against any type consistent with the
type the expression actually synthesizes.

\begin{thm}[Type Assignment Unicity]
  If $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$
  and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$
  then $\htau=\htau'$.
\end{thm}
\noindent
Consequently, analytic expansion reports the type actually assigned to the expansion it produces.
%
For example, we can derive that $\expandAna{\hGamma}{c}{\tehole}{c}{b}{\emptyset}$.% where $\emptyset$ is the empty hole context.

@@

When the programmer interactively fills a program hole, that program's
reduction and result are generally each affected, and the system
should respond by (quickly) updating the program's output.
%
Na\"ively, this response requires recomputing the program output
``from scratch'' after each such edit.
%
However, by virtue of its design principles (based on
CMTT), \HazelnutLive offers a more efficient approach: To \emph{resume
evaluation} of previously indeterminate expressions, which are
(generally) more determined after the programmer fills a hole than
beforehand.
%
We call this feature \emph{fill-and-resume}, by analogy to the
``edit-and-resume'' features available in some systems today, such as
Visual Studio (see below for a comparison)\todo{cite and compare}{}.

%%%%%%%%% New par
Formally, 
%
we interpret the hole environments as \emph{delayed substitutions},
the same interpretation suggested for closures in contextual modal
type theory (CMTT) by \citet{Nanevski2008}.
%
%In practice, it may be useful to cache results from several previous
%expansions, e.g., by employing off-the-shelf programming language
%abstractions for incremental computation~\cite{Hammer14,Hammer15}.
%
We define hole filling in terms of the expanded (internal) program,
and define filling a hole in an external language term as filling the
corresponding hole in its expansion.

>>>>

As with the type system for the external language (above), 
we specify expansion bidirectionally \cite{DBLP:conf/ppdp/FerreiraP14}.
%
The synthetic expansion judgement~$\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$ synthesizes a type~$\htau$ from~$\hexp$.
%
Synthetic expansion produces an internal expression~$d$ and a hole context~$\hDelta$.
%
%We say more about hole contexts, which are used in the type assignment judgement, $\hasType{\Delta}{\hGamma}{d}{\htau}$, below.
We describe hole contexts, which serve as ``inputs'' to the type assignment judgement~$\hasType{\Delta}{\hGamma}{d}{\htau}$, further below. 
%
The analytic expansion judgement~$\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$, checks~$\hexp$ against~$\htau$ and produces an expansion~$d$ of type~$\htau'$, and a hole context~$\hDelta$.
%
The following theorem establishes that expansions are well-typed and in the analytic case that the type $\htau'$ is consistent with $\htau$.
%
\begin{thm}[Typed Expansion]\label{thm:typed-expansion} ~
  \begin{enumerate}[nolistsep]
    \item
      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
    \item
      If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      then $\tconsistent{\htau}{\htau'}$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$.
  \end{enumerate}
\end{thm}
\noindent
%
%The reason analytic expansion produces an expansion of consistent
%type is because the subsumption rule, as previously discussed, allows
%us to check an external expression against any type consistent with
%the type the expression actually synthesizes, whereas every internal
%expression can be assigned at most one type, i.e. the following
%standard unicity property holds of the type assignment system.
%
Analytic expansion produces an expansion of consistent type:
%
Intuitively, the subsumption rule permits us
to check an external expression against any type consistent with the
type that the expression \emph{actually} synthesizes:
%
\begin{thm}[Type Assignment Unicity]
  If $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$
  and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$
  then $\htau=\htau'$.
\end{thm}
\noindent
Consequently, analytic expansion reports the type actually assigned to the expansion it produces.
%
For example, we can derive that $\expandAna{\hGamma}{c}{\tehole}{c}{b}{\emptyset}$.% where $\emptyset$ is the empty hole context.

@@

When the programmer interactively fills a program hole, that program's
reduction and result are generally each affected, and the system
should respond by (quickly) updating the program's output.
%
Na\"ively, this response requires recomputing the program output
``from scratch'' after each such edit.
%
However, by virtue of its design principles (based on
CMTT), \HazelnutLive offers a more efficient approach: To \emph{resume
evaluation} of previously indeterminate expressions, which are
(generally) more determined after the programmer fills a hole than
beforehand.
%
This \emph{fill-and-resume} behavior formalizes the (ad hoc)
``edit-and-resume'' behavior permitted by some current systems,
e.g.,~Microsoft Visual Studio.\todo{cite}

%(see below for a comparison)\todo{cite and compare}{}. ---NO TIME/SPACE for this.

%%%%%%%%% New par
Formally, 
%
we interpret hole environments as \emph{delayed substitutions},
the same interpretation suggested for closures in contextual modal
type theory (CMTT) by \citet{Nanevski2008}.
%
%In practice, it may be useful to cache results from several previous
%expansions, e.g., by employing off-the-shelf programming language
%abstractions for incremental computation~\cite{Hammer14,Hammer15}.
%
We define hole filling in terms of the expanded (internal) program,
and define filling a hole in an external language term as filling the
corresponding hole in its expansion.
